The promise of community-driven preprints in ecology and evolution

### 2024-10-15 ms/ms.qmd
Adding link to supplement
Added Table S1 ref in main MS for addressing C5

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has become a popular preprint server for ecologists and evolutionary biologists. The server has since been adopted by the California Digital Library (CDL). Editors are ecologists and evolutionary biologists from across the globe who volunteer their time to screen papers and push new initiatives in the preprint space. To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the accepted articles available on *EcoEvoRxiv* as of  `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). We consider both preprints and postprints as ‘articles’. After removing five duplicate titles – suggesting that a few authors uploaded the article as separate submissions rather than updating the existing article – we had data for a total of `r pr` articles with ~55--60 posted approximately monthly in the last two years ([@fig-summary]A).

@@

# Dataframes
            submissions <- data2          # Cntains all submissions, preprints, postprints and all article types
              preprints <- data2  %>% filter(time_between_preprint_and_pub_days >= 60) # Contains only preprints and not postprints
preprints_research_only <- data2  %>% filter(time_between_preprint_and_pub_days >= 60 & type_of_preprint == "research article") # Limits preprints to research only articles

  # Subset the preprints
  dates_preprints <- range(preprints$preprint_published_date)
##---------------------##
# Submissions to  EcoEvoRxiv over time (everything)
##---------------------##
  angle = 45
  p1 <- ggplot(data2, aes(x = preprint_pub_date)) +
    geom_histogram(fill = "white", color = "black") +
    scale_y_continuous(limits = c(0, 130), expand = c(0, 0)) +
    labs(x = "Published date", y = "Number of papers") +
    theme_classic() +
    #theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    theme(plot.tag = element_text(size = 16, face = "bold")) + 
      # Soft Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-06-01"),  xend = as.POSIXct("2018-06-01"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2018-06-01"), y = 85, label = "Soft Launch", size = 4, angle = angle, hjust = 0) +
    # Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-11-11"),  xend = as.POSIXct("2018-11-11"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2018-11-11"), y = 85, label = "Launch", size = 4, angle = angle, hjust = 0) +
    # Spanish and Port
    annotate( "segment", x = as.POSIXct("2023-02-14"),  xend = as.POSIXct("2023-02-14"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2022-11-14"), y = 85, label = "Spanish and Portug.", size = 4, angle = angle, hjust = 0) +
    # Transition to CDL
    annotate( "segment", x = as.POSIXct("2022-09-14"),  xend = as.POSIXct("2022-09-14"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2022-07-14"), y = 85, label = "CDL", size = 4, angle = angle, hjust = 0) +
   # IUCN Red Eco submission 29 August 2023
   # 1000 preprints
    annotate( "segment", x = as.POSIXct("2023-08-29"),  xend = as.POSIXct("2023-08-29"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2023-04-29"), y = 85, label = "IUCN Report", size = 4, angle = angle, hjust = 0)

@@

@@

We estimated how long it takes to publish a preprint in ecology and evolution by recording the time between when a preprint was first posted on *EcoEvoRxiv*, and its final acceptance in a peer-reviewed journal. In total, `r unpub` submissions remained unpublished (`r (unpub / nrow(data2))*100`%, n = `r nrow(data2)`) at the time when these data were collected. Not all of these preprints, however, are anticipated to be published in a peer-reviewed journal (e.g., reports). Nonetheless, the median time to publication for preprints was `r sum_preprints$median` days (8 months) (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days) with the maximum time to publication being `r sum_preprints$max` days or `r sum_preprints$max/365` years ([@fig-pubsummary]A). Our results largely confirm the extended timeframes that most authors experience between writing their research papers and their publication.

>>>>

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has become a popular preprint server for ecologists and evolutionary biologists. The server has since been adopted by the California Digital Library (CDL). Editors are ecologists and evolutionary biologists from across the globe who volunteer their time to screen papers and push new initiatives in the preprint space. To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the accepted articles available on *EcoEvoRxiv* as of  `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). We consider both preprints and postprints as ‘articles’. After removing five duplicate titles – suggesting that a few authors uploaded the article as separate submissions rather than updating the existing article – we had data for a total of `r pr` articles with ~55--60 posted approximately monthly in the last two years ([@fig-summary]A). For more details on the data collection process, see the Supplement (https://daniel1noble.github.io/ecoevo_1000/).

@@

# Dataframes
              preprints <- data2  %>% filter(time_between_preprint_and_pub_days >= 60) # Contains only preprints and not postprints
     preprints_pub_only <- data2  %>% filter(time_between_preprint_and_pub_days >= 60 & !is.na(preprint_published_date))

# Some descriptions

  dates_preprints <- range(preprints$preprint_published_date)

##---------------------##
# Submissions to  EcoEvoRxiv over time (everything)
##---------------------##
  angle = 45
  p1 <- ggplot(data2, aes(x = preprint_pub_date)) +
    geom_histogram(fill = "white", color = "black") +
    scale_y_continuous(limits = c(0, 130), expand = c(0, 0)) +
    labs(x = "Published date", y = "Number of papers") +
    theme_classic() +
    #theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    theme(plot.tag = element_text(size = 16, face = "bold")) + 
      # Soft Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-06-01"),  xend = as.POSIXct("2018-06-01"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2018-06-01"), y = 85, label = "Soft Launch", size = 4, angle = angle, hjust = 0) +
    # Launch of EcoEvoRxiv
    annotate( "segment", x = as.POSIXct("2018-11-11"),  xend = as.POSIXct("2018-11-11"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2018-11-11"), y = 85, label = "Launch", size = 4, angle = angle, hjust = 0) +
    # Spanish and Port
    annotate( "segment", x = as.POSIXct("2023-02-14"),  xend = as.POSIXct("2023-02-14"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2022-11-14"), y = 85, label = "Spanish and Portug.", size = 4, angle = angle, hjust = 0) +
    # Transition to CDL
    annotate( "segment", x = as.POSIXct("2022-09-14"),  xend = as.POSIXct("2022-09-14"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2022-07-14"), y = 85, label = "CDL", size = 4, angle = angle, hjust = 0) +
   # IUCN Red Eco submission 29 August 2023
   # 1000 preprints
    annotate( "segment", x = as.POSIXct("2023-08-29"),  xend = as.POSIXct("2023-08-29"), y = 0, yend = 82, color = "black", linetype = "dashed", size = 0.5) + 
    annotate("text", x = as.POSIXct("2023-04-29"), y = 85, label = "IUCN Report", size = 4, angle = angle, hjust = 0)

@@

# Publication time by article type
  preprints_pub_only_stats <- preprints_pub_only                                                                    %>% 
                              group_by(type_of_preprint)                                                            %>% 
                              summarise(mean_time = mean(time_between_preprint_and_pub_days, na.rm = TRUE), 
                                          sd_time = sd(time_between_preprint_and_pub_days, na.rm = TRUE), 
                                                n = n())                                                             %>% 
                              filter(!type_of_preprint == "report")                                                   %>% 
                              mutate(mean_time = round(mean_time, 2), sd_time = round(sd_time, 2))
  write.csv(preprints_pub_only_stats, here("output", "tables", "preprints_pub_only_stats.csv"))

@@

We estimated how long it takes to publish a preprint in ecology and evolution by recording the time between when a preprint was first posted on *EcoEvoRxiv*, and its final acceptance in a peer-reviewed journal. In total, `r unpub` submissions remained unpublished (`r (unpub / nrow(data2))*100`%, n = `r nrow(data2)`) at the time when these data were collected. Not all of these preprints, however, are anticipated to be published in a peer-reviewed journal (e.g., reports). Nonetheless, the median time to publication for preprints was `r sum_preprints$median` days (8 months) for all preprints that ended up being published (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days) with the maximum time to publication being `r sum_preprints$max` days or `r sum_preprints$max/365` years ([@fig-pubsummary]A). For a full breakdown on the time to publication based on article type see Supplement (Table S1). Our results largely confirm the extended timeframes that most authors experience between writing their research papers and their publication.
